Re-analysing the data from Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}): What can we learn from an under-powered absence of difference?

### 2022-02-11 manuscript.tex
minor updates
manucript second revision

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  english,
  man, donotrepeattitle,mask,floatsintext]{apa6}
\title{Re-analysing the data from \protect\hyperlink{ref-moffatt_inner_2020}{Moffatt et al.} (\protect\hyperlink{ref-moffatt_inner_2020}{2020}): What can we learn from an under-powered absence of difference?}
\author{Ladislas Nalborczyk\textsuperscript{1, 2, 3}}
\date{}

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Re-analysing the data from Moffatt et al. (2020): What can we learn from an under-powered absence of difference?},
  pdfauthor={Ladislas Nalborczyk1, 2, 3},
  pdflang={en-EN},
  pdfkeywords={NHST, Bayes factor, reanalysis, inner speech, rumination, electromyography},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

% Create new environments so endfloat can handle them
% \newenvironment{ltable}
%   {\begin{landscape}\centering\begin{threeparttable}}
%   {\end{threeparttable}\end{landscape}}
\newenvironment{lltable}{\begin{landscape}\centering\begin{ThreePartTable}}{\end{ThreePartTable}\end{landscape}}

@@

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\HyOrg@maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother
\shorttitle{The absence of evidence fallacy}
\keywords{NHST, Bayes factor, reanalysis, inner speech, rumination, electromyography}
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\ifXeTeX
  % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
  \usepackage{polyglossia}
  \setmainlanguage[]{english}
\else
  \usepackage[main=english]{babel}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
\fi
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

@@

\abstract{
\protect\hyperlink{ref-moffatt_inner_2020}{Moffatt et al.} (\protect\hyperlink{ref-moffatt_inner_2020}{2020}) reported the results of an experiment (N = 26 in the final sample) comparing the facial electromyographic correlates of mental rumination and distraction, following an experimentally induced stressor. Based on the absence of significant difference in the perioral muscular activity between the rumination and distraction conditions, \protect\hyperlink{ref-moffatt_inner_2020}{Moffatt et al.} (\protect\hyperlink{ref-moffatt_inner_2020}{2020}) concluded that \emph{self-reported} inner experience was unrelated to peripheral muscular activity as assessed using surface electromyography. We suggest this conclusion is at best hasty. Indeed, concluding on the absence of an effect based on an under-powered non-significant \emph{p}-value is strongly uninformative. In this short commentary, we show that there is limited evidence for the main conclusion put forward by \protect\hyperlink{ref-moffatt_inner_2020}{Moffatt et al.} (\protect\hyperlink{ref-moffatt_inner_2020}{2020}). We suggest ways forward, both from a theoretical and methodological perspective. Complete source code, reproducible analyses, and figures are available at \url{https://osf.io/ba3gk/}.
}

@@

Wordcount (excluding abstract, references, tables, and figures): 3168

\newpage

@@

Given the predominantly verbal nature of rumination (e.g., \protect\hyperlink{ref-ehring_repetitive_2008}{Ehring \& Watkins, 2008}; \protect\hyperlink{ref-goldwin_concreteness_2013}{Goldwin et al., 2013}; \protect\hyperlink{ref-goldwin_concreteness_2012}{Goldwin \& Behar, 2012}; \protect\hyperlink{ref-mclaughlin_effects_2007}{McLaughlin et al., 2007}), we previously proposed to consider rumination as a form of inner speech and to study it using the methods that have historically been used to study other forms of inner speech, namely, by using surface electromyography (EMG) and motor interference protocols (e.g., \protect\hyperlink{ref-nalborczyk_dissociating_2021}{Nalborczyk, Banjac, et al., 2021}; \protect\hyperlink{ref-nalborczyk_orofacial_2017}{Nalborczyk et al., 2017}, \protect\hyperlink{ref-nalborczyk_articulatory_2020}{2020}; \protect\hyperlink{ref-nalborczyk_understanding_2019}{Nalborczyk, 2019}). We first showed that induced rumination was accompanied by increased facial (both over a forehead and a perioral site) muscular activity in comparison to a rest period (\protect\hyperlink{ref-nalborczyk_orofacial_2017}{Nalborczyk et al., 2017}). However, because rumination was only compared to a rest period, it remained uncertain whether this perioral activity was specifically related to (inner) speech processes. Therefore, we ran a follow-up study comparing verbal to non-verbal rumination, which suggested that the facial EMG correlates we had previously identified were not specifically related to the verbal content of the ruminative thoughts (\protect\hyperlink{ref-nalborczyk_dissociating_2021}{Nalborczyk, Banjac, et al., 2021}). We discussed these findings in length and proposed several theoretical interpretations that can account for these results in the discussion of \protect\hyperlink{ref-nalborczyk_dissociating_2021}{Nalborczyk, Banjac, et al.} (\protect\hyperlink{ref-nalborczyk_dissociating_2021}{2021}) and more extensively in \protect\hyperlink{ref-nalborczyk_understanding_2019}{Nalborczyk} (\protect\hyperlink{ref-nalborczyk_understanding_2019}{2019}).

\protect\hyperlink{ref-moffatt_inner_2020}{Moffatt et al.} (\protect\hyperlink{ref-moffatt_inner_2020}{2020}) designed an experiment with the aim of refining our understanding of the involvement of the speech motor system in different varieties of inner speech and clarifying the relation between the peripheral correlates of inner speech and (self-reported) subjective experience. Their main conclusion is that inner experience between induced rumination and distraction differs ``without a change in electromyographic correlates of inner speech'' (p.1). In other words, they suggest that the subjective experience of inner speech is unrelated (or loosely related) to the electromyographic correlates of inner speech, which are thought to be represented mostly by the EMG amplitude recorded over the orbicularis oris inferior and orbicularis oris superior muscles. However, for this in-sample observation to be of interest in an out-of-sample context (i.e., to be informative for other non-observed individuals, or said otherwise, to bring information about the population), this absence of difference should be substantiated by adequately-powered statistical tests (given the target effect size) as well as reliable measures. This is unlikely to be the case here, for reasons that we will present and discuss in the present article.

\hypertarget{re-analysing-the-data-from-moffatt_inner_2020}{%
\section{\texorpdfstring{Re-analysing the data from \protect\hyperlink{ref-moffatt_inner_2020}{Moffatt et al.} (\protect\hyperlink{ref-moffatt_inner_2020}{2020})}{Re-analysing the data from Moffatt et al. (2020)}}\label{re-analysing-the-data-from-moffatt_inner_2020}}

@@

As typical in studies manipulating induced rumination, \protect\hyperlink{ref-moffatt_inner_2020}{Moffatt et al.} (\protect\hyperlink{ref-moffatt_inner_2020}{2020}) designed a two-step protocol. First, they aimed to induce a negative mood by asking participants to solve unsolvable or excessively difficult anagram and subtraction tasks. Second, they prompted participants to either ruminate on these (purportedly induced) negative feelings (by asking them to ``think about the causes, consequences, and meaning of their current feelings'') or to distract themselves (by asking them to ``think about a village, city or town that you are particularly familiar with''). Rumination and distraction were manipulated within-subject, with all subjects alternating between rumination and distraction, in a counter-balanced order.

Their final sample of participants, after data exclusion, included 26 participants (data available at \url{https://osf.io/hj7tz/}). The EMG data is depicted in Figure \ref{fig:general} by condition (where BAS, DIS, and RUM refer to the baseline, distraction, and rumination conditions, respectively) and by muscle (frontalis, FRO; orbicularis oris inferior, OOI; and orbicularis oris superior, OOS). This figure shows that the average natural logarithm of the EMG peak amplitude recorded over the FRO was at similar levels in the baseline and distraction conditions, but was much higher in the rumination condition. However, the average natural logarithm of the EMG peak amplitude recorded over the OOI and OOS muscles was higher than baseline in both the rumination and distraction conditions, with a slight increase from distraction to rumination (both on the mean and median). Having described the data collected by \protect\hyperlink{ref-moffatt_inner_2020}{Moffatt et al.} (\protect\hyperlink{ref-moffatt_inner_2020}{2020}), we now turn to a discussion of some problems related to conclusions that can be made from under-powered non-significant results.

@@

There is an infamous tradition of conducting and interpreting uninformative null-hypothesis significance tests in Psychology (e.g., \protect\hyperlink{ref-meehl_appraising_1990-1}{Meehl, 1990a}, \protect\hyperlink{ref-meehl_theory-testing_1967}{1967}, \protect\hyperlink{ref-meehl_theoretical_1978}{1978}, \protect\hyperlink{ref-meehl_why_1990}{1990b}, \protect\hyperlink{ref-harlow_problem_1997}{1997}). By ``uninformative,'' we mean that some null-hypothesis significance tests are simply not diagnostic with regards to the substantive effect of interest (e.g., whether there is a difference between conditions A and B).

As highlighted by several authors (e.g., \protect\hyperlink{ref-cohen_earth_1994}{J. Cohen, 1994}; \protect\hyperlink{ref-pollard_probability_1987}{Pollard \& Richardson, 1987}; \protect\hyperlink{ref-rouder_is_2016}{Rouder et al., 2016}), concluding that an effect is probably absent solely based on a non-significant \emph{p}-value is the continuous (i.e., probabilistic) extension of the modus tollens and is not a valid argument (i.e., the conclusion does not follow from the premises). This fallacious argument is also known as the \emph{fallacy of acceptance}, the \emph{absence of evidence fallacy} or \emph{the argument from ignorance}, and proceeds as follows: ``If the null hypothesis is true, then this observation should \emph{rarely} occur. This observation occurred. Therefore, the null hypothesis is false (or has low probability).'' In short, this argument is fallacious because it fails to consider the (probability of the data under the) alternative hypothesis.

@@

\hypertarget{enthusiastic-prior-specifications}{%
\subsection{Enthusiastic prior specifications}\label{enthusiastic-prior-specifications}}

Anticipating the legitimate critiques on the power of their study, \protect\hyperlink{ref-moffatt_inner_2020}{Moffatt et al.} (\protect\hyperlink{ref-moffatt_inner_2020}{2020}) report the results of a power analysis using the effect size reported in \protect\hyperlink{ref-nalborczyk_orofacial_2017}{Nalborczyk et al.} (\protect\hyperlink{ref-nalborczyk_orofacial_2017}{2017}) of \(d = 0.72\). This represents a highly optimistic estimate of the substantive effect of interest (i.e., the difference in the natural logarithm of the EMG peak amplitude between the rumination and distraction conditions) as this effect represents the standardised mean difference in EMG amplitude \emph{between a rest and a rumination periods} as estimated in \protect\hyperlink{ref-nalborczyk_orofacial_2017}{Nalborczyk et al.} (\protect\hyperlink{ref-nalborczyk_orofacial_2017}{2017}).

@@

We suggest the (a priori) power of the study ran by \protect\hyperlink{ref-moffatt_inner_2020}{Moffatt et al.} (\protect\hyperlink{ref-moffatt_inner_2020}{2020}) was much lower than suggested by the authors. Indeed, we speculate that the standardised mean difference in EMG peak amplitude between the rumination and distraction conditions may be much weaker than the standardised mean difference in EMG amplitude between the rumination and rest conditions. If we assume that the former is half the size of the latter, therefore the a priori power of the main statistical test from \protect\hyperlink{ref-moffatt_inner_2020}{Moffatt et al.} (\protect\hyperlink{ref-moffatt_inner_2020}{2020}) was around \(0.42\), meaning that they had less than 1 chance out of 2 to find a significant effect (given that the population effect size was actually \(0.36\)). Notice that whereas taking half the effect size of \protect\hyperlink{ref-nalborczyk_orofacial_2017}{Nalborczyk et al.} (\protect\hyperlink{ref-nalborczyk_orofacial_2017}{2017}) may seem arbitrary, Figure \ref{fig:power} shows that a one-sample t-test with a sample size of \(N = 26\) is under-powered for a vast range of effect sizes.

@@

Once again, anticipating the legitimate critique that the absence of a significant difference is not necessarily ``significant'' evidence for the absence of an effect, \protect\hyperlink{ref-moffatt_inner_2020}{Moffatt et al.} (\protect\hyperlink{ref-moffatt_inner_2020}{2020}) reported the following Bayes factor (BF) analysis (p.12):

@@

\caption{Illustrating the distribution of Bayes factors in favour of the alternative hypothesis for different population effect sizes (N = 26). In the left panel, the effect size is fixed to d = 0 (i.e., the null hypothesis), in the middle panel, it is fixed to d = 0.36 (i.e., the supposed target effect size in Moffatt et al., 2020), and in the right panel, the effect size is fixed to d = 0.72 (i.e., the effect size reported in Nalborczyk et al., 2017). The red vertical dashed line indicates the value of the BF computed for the OOI by Moffatt et al. (2020), on the log scale.}\label{fig:bf-dance}
\end{figure}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  man, donotrepeattitle,floatsintext]{apa6}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Re-analysing the data from Moffatt et al. (2020): What can we learn from an under-powered absence of difference?},
  pdfauthor={Ladislas Nalborczyk1, 2, 3},
  pdflang={en-EN},
  pdfkeywords={NHST, Bayes factor, reanalysis, inner speech, rumination, electromyography},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

% Create new environments so endfloat can handle them
% \newenvironment{ltable}
%   {\begin{landscape}\begin{center}\begin{threeparttable}}
%   {\end{threeparttable}\end{center}\end{landscape}}
\newenvironment{lltable}{\begin{landscape}\begin{center}\begin{ThreePartTable}}{\end{ThreePartTable}\end{center}\end{landscape}}

@@

% Overwrite redefinition of paragraph and subparagraph by the default LaTeX template
% See https://github.com/crsh/papaja/issues/292
\makeatletter
\renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-1em}%
  {\normalfont\normalsize\bfseries\itshape\typesectitle}}

\renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-\z@\relax}%
  {\normalfont\normalsize\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
\makeatother

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\HyOrg@maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{NHST, Bayes factor, reanalysis, inner speech, rumination, electromyography\newline\indent Word count: 3152 (excluding abstract, references, tables, and figures)}
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\title{Re-analysing the data from Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}): What can we learn from an under-powered absence of difference?}
\author{Ladislas Nalborczyk\textsuperscript{1, 2, 3}}
\date{}

\shorttitle{The absence of evidence fallacy}

\authornote{

Correspondence concerning this article should be addressed to Ladislas Nalborczyk, LPC, CNRS, Aix-Marseille University, 13 Boulevard Gustave Desplaces, 13003 Marseille, France. E-mail: \href{mailto:ladislas.nalborczyk@univ-amu.fr}{\nolinkurl{ladislas.nalborczyk@univ-amu.fr}}

@@

\abstract{%
Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}) reported the results of an experiment (N = 26 in the final sample) comparing the facial electromyographic correlates of mental rumination and distraction, following an experimentally induced stressor. Based on the absence of significant difference (and BFs between 3.6 and 4.3) in the perioral muscular activity between the rumination and distraction conditions, Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}) concluded that \emph{self-reported} inner experience was unrelated to peripheral muscular activity as assessed using surface electromyography. In this short commlentary, we show that there is limited evidence for the main conclusion put forward by Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}), and we suggest ways forward, both from a theoretical and a methodological perspective. Complete source code, reproducible analyses, and figures are available at \url{https://osf.io/ba3gk/}.
}

@@

@@

Given the predominantly verbal nature of rumination (e.g., \protect\hyperlink{ref-ehring_repetitive_2008}{Ehring \& Watkins, 2008}; \protect\hyperlink{ref-goldwin_concreteness_2013}{Goldwin et al., 2013}; \protect\hyperlink{ref-goldwin_concreteness_2012}{Goldwin \& Behar, 2012}; \protect\hyperlink{ref-mclaughlin_effects_2007}{McLaughlin et al., 2007}), we previously proposed to consider rumination as a form of inner speech and to study it using the methods that have historically been used to study other forms of inner speech, namely, by using surface electromyography (EMG) and motor interference protocols (e.g., \protect\hyperlink{ref-nalborczyk_dissociating_2021}{Nalborczyk, Banjac, et al., 2021}; \protect\hyperlink{ref-nalborczyk_articulatory_2022}{Nalborczyk et al., 2022}, \protect\hyperlink{ref-nalborczyk_orofacial_2017}{2017}; \protect\hyperlink{ref-nalborczyk_understanding_2019}{Nalborczyk, 2019}). We first showed that induced rumination was accompanied by increased facial (both over a forehead and a perioral site) muscular activity in comparison to a rest period (\protect\hyperlink{ref-nalborczyk_orofacial_2017}{Nalborczyk et al., 2017}). However, because rumination was only compared to a rest period, it remained uncertain whether this perioral activity was specifically related to (inner) speech processes. Therefore, we ran a follow-up study comparing verbal to non-verbal rumination, which suggested that the facial EMG correlates we had previously identified were not specifically related to the verbal content of the ruminative thoughts (\protect\hyperlink{ref-nalborczyk_dissociating_2021}{Nalborczyk, Banjac, et al., 2021}). We discussed these findings in length and proposed several theoretical interpretations that can account for these results in the discussion of Nalborczyk, Banjac, et al. (\protect\hyperlink{ref-nalborczyk_dissociating_2021}{2021}) and more extensively in Nalborczyk (\protect\hyperlink{ref-nalborczyk_understanding_2019}{2019}).

Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}) designed an experiment with the aim of refining our understanding of the involvement of the speech motor system in different varieties of inner speech and clarifying the relation between the peripheral correlates of inner speech and (self-reported) subjective experience. Their main conclusion is that inner experience between induced rumination and distraction differs ``without a change in electromyographic correlates of inner speech'' (p.1). In other words, they suggest that the subjective experience of inner speech is unrelated (or loosely related) to the electromyographic correlates of inner speech, which are thought to be represented by the EMG amplitude recorded over the orbicularis oris inferior and orbicularis oris superior muscles. However, for this in-sample observation to be of interest in an out-of-sample context (i.e., to be informative for other non-observed individuals, or said otherwise, to bring information about the population), this absence of difference should be substantiated by adequately-powered statistical tests (given the target effect size) as well as reliable measures. This is unlikely to be the case here, for reasons that we will present and discuss in the present article.

\hypertarget{reconsidering-the-data-from-moffatt_inner_2020}{%
\section{\texorpdfstring{Reconsidering the data from Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020})}{Reconsidering the data from Moffatt et al. (2020)}}\label{reconsidering-the-data-from-moffatt_inner_2020}}

@@

As typical in studies manipulating induced rumination, Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}) designed a two-step protocol. First, they aimed to induce a negative mood by asking participants to solve unsolvable or excessively difficult anagram and subtraction tasks. Second, they prompted participants to either ruminate on these (purportedly induced) negative feelings (by asking them to ``think about the causes, consequences, and meaning of their current feelings'') or to distract themselves (by asking them to ``think about a village, city or town that you are particularly familiar with''). Rumination and distraction were manipulated within-subject, with all subjects alternating between rumination and distraction, in a counter-balanced order.

Their final sample of participants, after data exclusion, included 26 participants (data available at \url{https://osf.io/hj7tz/}). The EMG data is depicted in Figure \ref{fig:general} by condition (where BAS, DIS, and RUM refer to the baseline, distraction, and rumination conditions, respectively) and by muscle (frontalis, FRO; orbicularis oris inferior, OOI; and orbicularis oris superior, OOS). This figure shows that the average natural logarithm of the EMG peak amplitude recorded over the FRO was at similar levels in the baseline and distraction conditions, but was much higher in the rumination condition. However, the average natural logarithm of the EMG peak amplitude recorded over the OOI and OOS muscles was higher than baseline in both the rumination and distraction conditions, with a slight increase from distraction to rumination (both on the mean and median). Having described the data collected by Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}), we now turn to a discussion of some problems related to conclusions that can be made from under-powered non-significant results.

@@

There is an infamous tradition of conducting and interpreting uninformative null-hypothesis significance tests in Psychology (e.g., \protect\hyperlink{ref-meehl_appraising_1990-1}{Meehl, 1990a}, \protect\hyperlink{ref-meehl_theory-testing_1967}{1967}, \protect\hyperlink{ref-meehl_theoretical_1978}{1978}, \protect\hyperlink{ref-meehl_why_1990}{1990b}, \protect\hyperlink{ref-harlow_problem_1997}{1997}). By ``uninformative'', we mean that some null-hypothesis significance tests are simply not diagnostic with regards to the substantive effect of interest (e.g., whether there is a difference between conditions A and B).

As highlighted by several authors (e.g., \protect\hyperlink{ref-cohen_earth_1994}{J. Cohen, 1994}; \protect\hyperlink{ref-pollard_probability_1987}{Pollard \& Richardson, 1987}; \protect\hyperlink{ref-rouder_is_2016}{Rouder et al., 2016}), concluding that an effect is probably absent solely based on a non-significant \emph{p}-value is the continuous (i.e., probabilistic) extension of the modus tollens and is not a valid argument (i.e., the conclusion does not follow from the premises). This fallacious argument is also known as the \emph{fallacy of acceptance}, the \emph{absence of evidence fallacy} or \emph{the argument from ignorance}, and proceeds as follows: ``If the null hypothesis is true, then this observation should \emph{rarely} occur. This observation occurred. Therefore, the null hypothesis is false (or has low probability)''. In short, this argument is fallacious because it fails to consider the (probability of the data under the) alternative hypothesis.

@@

\hypertarget{optimistic-a-priori-power-analysis}{%
\subsection{Optimistic a priori power analysis}\label{optimistic-a-priori-power-analysis}}

Anticipating the legitimate critiques on the power of their study, Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}) report the results of a power analysis using the effect size reported in Nalborczyk et al. (\protect\hyperlink{ref-nalborczyk_orofacial_2017}{2017}) of \(d = 0.72\). This represents a highly optimistic estimate of the substantive effect of interest (i.e., the difference in the natural logarithm of the EMG peak amplitude between the rumination and distraction conditions) as this effect represents the standardised mean difference in EMG amplitude \emph{between a rest and a rumination periods} as estimated in Nalborczyk et al. (\protect\hyperlink{ref-nalborczyk_orofacial_2017}{2017}).

@@

We suggest the (a priori) power of the study ran by Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}) was much lower than suggested by the authors. Indeed, we speculate that the standardised mean difference in EMG peak amplitude between the rumination and distraction conditions may be much weaker than the standardised mean difference in EMG amplitude between the rumination and rest conditions. If we assume that the former is half the size of the latter, therefore the a priori power of the main statistical test from Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}) was around \(0.42\), meaning that they had less than 1 chance out of 2 to find a significant effect (given that the population effect size was actually \(0.36\)). Notice that whereas taking half the effect size of Nalborczyk et al. (\protect\hyperlink{ref-nalborczyk_orofacial_2017}{2017}) may seem arbitrary, Figure \ref{fig:power} shows that a one-sample t-test with a sample size of \(N = 26\) is under-powered for a vast range of effect sizes.

@@

Once again, anticipating the legitimate critique that the absence of a significant difference is not necessarily ``significant'' evidence for the absence of an effect, Moffatt et al. (\protect\hyperlink{ref-moffatt_inner_2020}{2020}) reported the following Bayes factor (BF) analysis (p.12):

@@

\caption{Illustrating the distribution of Bayes factors in favour of the alternative hypothesis for different population effect sizes (N = 26). In the left panel, the effect size is fixed to d = 0 (i.e., the null hypothesis), in the middle panel, it is fixed to d = 0.36 (i.e., the supposed target effect size in Moffatt et al., 2020), and in the right panel, the effect size is fixed to d = 0.72 (i.e., the effect size reported in Nalborczyk et al., 2017). The red vertical dashed line indicates the value of the BF computed for the OOI by Moffatt et al. (2020), on the logarithmic scale. The grey shaded area represents the conventional (but questionable) interval in which BFs are usually considered as inconclusive.}\label{fig:bf-dance}
\end{figure}
